---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 0
---

# Data: Basics

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

Additionally, we'll have our first look at the *pipe operator*, which will be
super useful for writing code that's readable.

*Reading*: (None)

```{r setup}
library(tidyverse)
```

```{r options, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

## First Checks
<!-- -------------------------------------------------- -->

### __q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

### __q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages, eval=FALSE}
?diamonds
```

The `cut` variable specifies the quality of a diamond's cut, while the `color` variable specifies the diamond's color and the `clarity` variable specifies how clear the diamond is.

### __q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?

```{r}
glimpse(diamonds)
```
`diamonds` includes the variables: carat, cut, color, clarity, depth, table, price, x, y, and z.

### __q3__ Run `summary(diamonds)`; what are the common values for each of the
variables? How widely do each of the variables vary?

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r q3-task}
summary(diamonds)
```

The average carat value is 0.7979, depth is 61.75, table value is 57.46, price is 3933, x is 5.731, y is 5.735, and z is 3.539. The most common cut rating is ideal, color is G, and clarity is SI1. The difference in count between the most and least common types of cuts is much higher than the differences for color and clarity. For the variables with numeric values, the values in price vary the most. The differences between the minimum and maximum value in the other variables are all less than 100, while the difference for price is over 18000.

**Observations**:

- There is a large variance in prices for diamonds
- Diamonds most commonly have a middle-of-the-road color
- Many diamonds have an "ideal" cut, while "fair" cuts are much rarer
- Most diamonds have a weight on the lower side
- Diamonds are more likely to be less clear
- The average length and width of diamonds are similar, but the maximum width is much larger than the maximum length

You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

## The Pipe Operator
<!-- -------------------------------------------------- -->

Throughout this class we're going to make heavy use of the *pipe operator*
`%>%`. This handy little function will help us make our code more readable.
Whenever you see `%>%`, you can translate that into the word "then". For
instance

```{r pipe-example}
diamonds %>%
  group_by(cut) %>%
  summarize(carat_mean = mean(carat))
```

Would translate into the tiny "story"

- Take the `diamonds` dataset, *then*
- Group it by the variable `cut`, *then*
- summarize it by computing the `mean` of `carat`

*What the pipe actually does*. The pipe operator `LHS %>% RHS` takes its
left-hand side (LHS) and inserts it as an the first argument to the function on
its right-hand side (RHS). So the pipe will let us take `glimpse(diamonds)` and
turn it into `diamonds %>% glimpse()`.

### __q4__ Use the pipe operator to re-write `summary(diamonds)`.

```{r q4-task}
diamonds %>% 
  summary()
```

## Reading Data
<!-- -------------------------------------------------- -->

So far we've only been looking at built-in datasets. Ultimately, we'll want to read in our own data. We'll get to the art of loading and *wrangling* data later, but for now, know that the `readr` package provides us tools to read data. Let's quickly practice loading data below.

### __q5__ Use the function `read_csv()` to load the file `"./data/tiny.csv"`.

```{r q5-task}
read_csv("./data/tiny.csv")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd).
