---
title: "RMS Titanic"
author: "Nerissa Theisen"
date: 2026-09-06
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

Variables:

- Class
- Sex
- Age
- Survived
- n

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Are there any differences?
  - The Wikipedia article states that the Titanic had 2208 people on board. However, the        total we computed was 2201. The difference between the totals is small, at 7 people.
  
- If yes, what might account for those differences?
  - The difference may be caused by unclear or partially lost physical logs of passengers      and crew members. For example, passengers may have missed or cancelled their trip and      not actually ended up on the ship, while their names were still kept in records. The       crew was also not a permanent crew and some of them may have not been properly             accounted for.

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Sex, y = n, fill = Class)) +
  geom_col(position = "dodge")
```

**Observations**:

- There were more male survivors than female survivors.
- The majority of female survivors were passengers, not crew.
- A little over half of the male survivors were crew, not passengers.
- There were many more male crew survivors than female crew survivors.
- For males, 3rd class had the highest number of survivors, while 2nd class had the lowest.
- For females, 1st class had the highest number of survivors, while 3rd class had the        lowest.
- 3rd class had a similar number of both male and female survivors, while there were many    more female survivors than male survivors for 1st and 2nd class.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
## Visualizes proportions against `Class` and `Sex`
df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup() %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Sex, y = Prop, fill = Class)) +
  geom_col(position = "dodge")
```

**Observations**:

- Write your observations here.
  - The last chart showed that fewer female crew members survived than male crew members.      However, this one reveals that most female crew members survived, while most male crew     members died, so there must have been less female crew members.
  - The last chart showed that more males survived than females, while this one shows that     a greater proportion of female 3rd class passengers survived than male 3rd class           passengers.
  - According to the plot, all female and male 1st and 2nd class passengers survived.
  - Females, whether 3rd class passengers or crew members, were more likely to survive than     male 3rd class passengers or crew members, which was not apparent in the last chart.
  
- Is there anything *fishy* going on in your plot?
  - Both the 1st and 2nd class survivors for males and females have a proportion of exactly     1, which is a very big coincidence.

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
## Visualizes proportions against `Class` and `Sex`, and `Age`
df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup() %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(x = Sex, y = Prop)) +
  geom_col() +
  facet_grid (Age ~ Class)
```

**Observations**:

- (Write your observations here.)
  - All child passengers in 1st and 2nd class survived, while less than half of both sexes     survived in 3rd class.
  - Almost all female passengers in 1st class survived.
  - Across the board, a greater proportion of females survived than males.
  - 1st class had the greatest proportion of survivors for both sexes.
  - For adult females, the proportion of survivors decreases with each class. For adult        males, 1st class has the highest proportion, 3rd class has the second highest              proportion, and 2nd class has the lowest proportion.
  - The group with the lowest proportion of survivors overall was adult males in 2nd class.

- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - The strange proportions for 1st and 2nd class survivors in the last bar chart were       likely caused by stacking. The new chart shows that all children in both classes           survived, while not all adults did. Since the old visual didn't take separations between   age groups into account, it ended up only showing the proportions for children. 

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
